---
title: "Counting number of transitions  & rate through time"
author: "Sridhar Halali"
date: "2023-10-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Importing required libraries
```{r setup, message=F}
library(phytools)
library(ape)
library(tidyverse)
library(conflicted)
library(viridis)
```

# Setting working directory
```{r setup, message=F}
setwd("/Users/sridhar/Library/CloudStorage/OneDrive-LundUniversity/LU box backup/Evolution of diapause/evo of diapause/Updated_diapause_analyses_19_sep_2023")
```

# Reading stochastic maps saved as .RDS files
```{r}
# Reading stochastic simulations RDS file if already not in the R environment
SimmapFiveState_flat <- readRDS("simmap_5state_1000sim_ARD_flat.RDS")
SimmapThreeState_flat <- readRDS("simmap_3state_1000sim_ARD_flat.RDS")
SimmapBinaryState_flat <- readRDS("simmap_Binary_1000sim_ARD_flat.RDS")

SimmapFiveState_fitz <- readRDS("simmap_5state_1000sim_ARD_fitzjohn.RDS")
SimmapThreeState_fitz <- readRDS("simmap_3state_1000sim_ARD_fitzjohn.RDS")
SimmapBinaryState_fitz <- readRDS("simmap_Binary_1000sim_ARD_fitzjohn.RDS")
```


# Counting number of transitions

##* Extracting number of transitions on each stochastic maps using countSimmap function
```{r}
# Flat prior
counts_FiveStates_flat <- data.frame(countSimmap(SimmapFiveState_flat)[[1]])
counts_ThreeStates_flat <- data.frame(countSimmap(SimmapThreeState_flat)[[1]])
counts_BinaryStates_flat <- data.frame(countSimmap(SimmapBinaryState_flat)[[1]])

# Fitzjohn prior
counts_FiveStates_fitz <- data.frame(countSimmap(SimmapFiveState_fitz)[[1]])
counts_ThreeStates_fitz <- data.frame(countSimmap(SimmapThreeState_fitz)[[1]])
counts_BinaryStates_fitz <- data.frame(countSimmap(SimmapBinaryState_fitz)[[1]])
```

##* Some data carpentry to get data into shape for plotting
```{r}
head(counts_FiveStates_fitz)

# Replacing '.' from the column names to -> & creating a new column showing root_prior

# Flat prior
names(counts_FiveStates_flat) <- gsub(x = names(counts_FiveStates_flat), pattern = "\\.", replacement = "->")  
counts_FiveStates_flat <- counts_FiveStates_flat %>% mutate(root_prior = "flat")

names(counts_ThreeStates_flat) <- gsub(x = names(counts_ThreeStates_flat), pattern = "\\.", replacement = "->")  
counts_ThreeStates_flat <- counts_ThreeStates_flat %>% mutate(root_prior = "flat")

names(counts_BinaryStates_flat) <- gsub(x = names(counts_BinaryStates_flat), pattern = "\\.", replacement = "->")  
counts_BinaryStates_flat <- counts_BinaryStates_flat %>% mutate(root_prior = "flat")


# Fitzjohn prior
names(counts_FiveStates_fitz) <- gsub(x = names(counts_FiveStates_fitz), pattern = "\\.", replacement = "->")  
counts_FiveStates_fitz <- counts_FiveStates_fitz %>% mutate(root_prior = "fitzjohn")

names(counts_ThreeStates_fitz) <- gsub(x = names(counts_ThreeStates_fitz), pattern = "\\.", replacement = "->")  
counts_ThreeStates_fitz <- counts_ThreeStates_fitz %>% mutate(root_prior = "fitzjohn")

names(counts_BinaryStates_fitz) <- gsub(x = names(counts_BinaryStates_fitz), pattern = "\\.", replacement = "->")  
counts_BinaryStates_fitz <- counts_BinaryStates_fitz %>% mutate(root_prior = "fitzjohn")


#============= Combining dataframes according to state classifications==================================#
# Five states
counts_FiveStates_final <- rbind(counts_FiveStates_flat, counts_FiveStates_fitz)

# Three states
counts_ThreeStates_final <- rbind(counts_ThreeStates_flat, counts_ThreeStates_fitz)

# Binary state
counts_BinaryStates_final <- rbind(counts_BinaryStates_flat, counts_BinaryStates_fitz)
#==================================================================================================#


#================= Converting wide to long data frame ==================================#
# Five states
counts_FiveStates_final <- counts_FiveStates_final %>% 
  select(-N) %>% 
  pivot_longer(!root_prior, names_to = "transitions", values_to = "count") %>% 
  mutate(transitions= factor(transitions),
         root_prior = factor(root_prior))

# Three states
counts_ThreeStates_final <- counts_ThreeStates_final %>% 
  select(-N) %>% 
  pivot_longer(!root_prior, names_to = "transitions", values_to = "count") %>% 
  mutate(transitions= factor(transitions),
         root_prior = factor(root_prior))

# Binary states
counts_BinaryStates_final <- counts_BinaryStates_final %>% 
  select(-N) %>% 
  pivot_longer(!root_prior, names_to = "transitions", values_to = "count") %>% 
  mutate(transitions= factor(transitions),
         root_prior = factor(root_prior))


#================= Removing same transitions ==================================#
counts_FiveStates_final <- counts_FiveStates_final %>% 
  dplyr::filter(!transitions %in% c("Adult->Adult", "Egg->Egg", "Larva->Larva", "Pupa->Pupa", "Flex->Flex"))

counts_ThreeStates_final <- counts_ThreeStates_final %>% 
  dplyr::filter(!transitions %in% c("Adult->Adult", "Juvenile->Juvenile", "Flex->Flex"))

counts_BinaryStates_final <- counts_BinaryStates_final %>% 
  dplyr::filter(!transitions %in% c("Flex->Flex", "Diapause->Diapause"))
```


##* Calculating average number of transitions
```{r}
# Five state
counts_FiveStates_average <- counts_FiveStates_final %>% 
  group_by(root_prior, transitions) %>% 
  summarise(mean_counts = mean(count))


# Three state
counts_ThreeStates_average <- counts_ThreeStates_final %>% 
  group_by(root_prior, transitions) %>% 
  summarise(mean_counts = mean(count))

# Binary state
counts_BinaryStates_average <- counts_BinaryStates_final %>% 
  group_by(root_prior, transitions) %>% 
  summarise(mean_counts = mean(count))
```


##* Plotting number of transitions
```{r}
# Setting the theme
theme_counts <- theme_bw() + 
  theme(panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())


# Figure 2a: Five state
pdf("transitions_FiveState.pdf", height=6, width=6)

counts_FiveStates_final %>% 
  ggplot(aes(x=count, y= reorder(transitions, count))) +
  
  geom_jitter(aes(fill=root_prior), height=0.2, alpha=0.1, shape=21, stroke=NA) +
  
  geom_jitter(data=counts_FiveStates_average,
             aes(x=mean_counts, y= transitions, fill=root_prior), 
             height = 0.15, shape=21, stroke=0.5, size=4) +
 
  scale_x_continuous(limits= c(-0.9, 250)) +
             
  scale_fill_manual(values = c("flat" = "grey", "fitzjohn" = "salmon")) +

  xlab("Number of transitions") + ylab("Transition between states") +
  theme_counts 

dev.off()


# Figure 2b: Three state
pdf("transitions_ThreeState.pdf", height=6, width=6)

counts_ThreeStates_final %>% 
  ggplot(aes(x=count, y= reorder(transitions, count))) +
  
  geom_jitter(aes(fill=root_prior), height=0.1, alpha=0.1, shape=21, stroke=NA) +
  
  geom_jitter(data=counts_ThreeStates_average,
             aes(x=mean_counts, y= transitions, fill=root_prior), 
             height = 0.09, shape=21, stroke=0.5, size=4) +
  
  scale_x_continuous(limits=c(-0.9, 150)) +
             
  scale_fill_manual(values = c("flat" = "grey", "fitzjohn" = "salmon")) +
  
  xlab("Number of transitions") + ylab("Transition between states") +
  theme_counts 

dev.off()


# Figure 2c: Binary state
pdf("transitions_BinaryState.pdf", height=6, width=6)

counts_BinaryStates_final %>% 
  ggplot(aes(x=count, y= reorder(transitions, count))) +
  
  geom_jitter(aes(fill=root_prior), height=0.07, alpha=0.1, shape=21, stroke=NA) +
  
  geom_jitter(data=counts_BinaryStates_average,
             aes(x=mean_counts, y= transitions, fill=root_prior), 
             height = 0.05, shape=21, stroke=0.5, size=4) +
             
  scale_fill_manual(values = c("flat" = "grey", "fitzjohn" = "salmon")) +
  
  xlab("Number of transitions") + ylab("Transition between states") +
  theme_counts 

dev.off()

```


# Calculating Rate Trough time
Calculating rate through time using 1000 stochastic maps. The rate is calculated by dividing the number transitions by the edge length in a given bin or segment. Here I am dividing the entire phylogeny into 30 bins (i.e. 107.60/30), thus, one bin equals to ~3.58 million years. Type="all" estimates total rates by taking all transitions into account. See Methods section for details. 

##* Loading functions for calculating and plotting rate through time. 
Only some functions are modified from the original code that can be found here (https://github.com/jakeberv/mammal_arboreality). The original getrates(), rateplot() and segplot() functions were especially modified to able to compute and plot confidence intervals around the rates. Function were modified by Etka Yapar. Please load both source codes. 

```{r}

# Loading original function (can be found here: https://github.com/jakeberv/mammal_arboreality)
# Original article can be found here:  https://doi.org/10.1002/ece3.8114
source("source_rate_through_time_Hughes_et_al_2021.R")

# Loading modified function (based on the original function mentioned above) for incorporating confidence bounds on rate through time calculations 
source("getRates_modified_Etka Yapar.R")


# Total age of the tree
ButterflyPhylo <- read.nexus("butterfly_phylogeny_pruned.trees")
max(nodeHeights(ButterflyPhylo)) # 107.60

```

##* Binary state diapause clasification

####- flat root prior
```{r}
# Calculating rates for each transition
counts.FD_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_flat, type="Flex->Diapause", uncert = "confint")

counts.DF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_flat, type="Diapause->Flex", uncert= "confint")


# Figure S7: Plotting rate through time for transitions for binary state diapause classification with flat root prior
pdf("newRate_BinaryState_flat.pdf", height=7, width=7)

rateplot_ey(rates=counts.DF_flat, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint = T)
par(new=T)
segplot_ey(rates=counts.FD_flat, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)
legend("topleft", 
       legend=c("diapause->flex", "flex->diapause"),     
       col = c("#440154FF", "#7AD151FF"), bty = 'n', lty=1, lwd=5)
       
dev.off()
```


####- fitzjohn root prior
```{r}
# Calculating rates for each transition
counts.DF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_fitz, type="Diapause->Flex", uncert="confint")
counts.FD_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_fitz, type="Flex->Diapause", uncert="confint")


# Figure 5c: Plotting rate through time for transitions for binary state diapause classification with fitzjohn root prior
pdf("newRate_BinaryState_fitz.pdf", height=7, width=7)

rateplot_ey(rates=counts.DF_fitz, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint = T)
par(new=T)
segplot_ey(rates=counts.FD_fitz, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)
legend("topleft", 
       legend=c("diapause->flex", "flex->diapause"),     
       col = c("#440154FF", "#7AD151FF"), bty = 'n', lty=1, lwd=5)
      
dev.off()
```


##* Three state dipauase classification

####- flat root prior 
```{r}
# Calculating rates for each transition
counts.AJ_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Adult->Juvenile", uncert = "confint")
counts.JA_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Juvenile->Adult", uncert = "confint")
counts.JF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Juvenile->Flex", uncert = "confint")
counts.FJ_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Flex->Juvenile", uncert = "confint")
counts.FA_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Flex->Adult", uncert = "confint")
counts.AF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Adult->Flex", uncert = "confint")

# Figure S7: Plotting rates
pdf("newRate_ThreeState_flat.pdf", height=7, width=7)

rateplot_ey(rates=counts.JF_flat, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint=T)

par(new=T)
segplot_ey(rates=counts.AJ_flat, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.JA_flat, ylim=c(0,0.02), color= "#2A788EFF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.FJ_flat, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.FA_flat, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.AF_flat, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=1, confint=T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("Juv->Flex", "Adult->Juv",  "Juv->Adult", "Flex->Juv", "Flex->Adult", "Adult->flex"),
       col = c("#482878FF", "#414487FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()
```


####- fitzohn root prior
```{r}
# Calculating rates for each transition
counts.AJ_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Adult->Juvenile", uncert = "confint")
counts.JA_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Juvenile->Adult", uncert = "confint")
counts.JF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Juvenile->Flex", uncert = "confint")
counts.FJ_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Flex->Juvenile", uncert = "confint")
counts.FA_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Flex->Adult", uncert = "confint")
counts.AF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Adult->Flex", uncert = "confint")

# Figure 5b: Plot rate through time
pdf("newRate_ThreeState_fitz.pdf", height=7, width=7)

rateplot_ey(rates=counts.JF_fitz, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint = T)

par(new=T)
segplot_ey(rates=counts.AJ_fitz, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.JA_fitz, ylim=c(0,0.02), color= "#2A788EFF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FJ_fitz, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FA_fitz, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.AF_fitz, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=1, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("Juv->Flex", "Adult->Juv",  "Juv->Adult", "Flex->Juv", "Flex->Adult", "Adult->flex"),
       col = c("#482878FF", "#414487FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()
```

##* Five state diapause classification
Note that for five state classification, rates are for only the high frequency transitions are calculated (see Figure 2a in the main text for number of transitions between states)

####- flat root prior
```{r}
# Calculating rates only for higher frequency transitions
counts.LF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Larva->Flex", uncert = "confint")
counts.FL_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Flex->Larva", uncert = "confint")
counts.PF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Pupa->Flex", uncert = "confint")
counts.PL_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Pupa->Larva", uncert = "confint")
counts.FP_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Flex->Pupa", uncert = "confint")
counts.FA_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Flex->Adult", uncert = "confint")


# Figure S7: Plotting rate through time for higher frequency shifts
pdf("newRate_FiveStates_flat.pdf", height=7, width=7)

rateplot_ey(rates=counts.LF_flat, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.1, lty=1, color = "#440154FF", confint = T)
         
par(new=T)
segplot_ey(rates=counts.FL_flat, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.PF_flat, ylim=c(0,0.02), color= "#2A788EFF" ,  spline=T, width=3, ax=F, lty=1, confint = T)

par(new=T)
segplot_ey(rates=counts.PL_flat, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty = 2, confint = T)

par(new=T)
segplot_ey(rates=counts.FP_flat, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FA_flat, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("Larva->Flex",  "Flex->Larva", "Pupa->Flex", "Pupa->Larva", "Flex->Pupa","Flex->Adult"),  
       col = c("#440154FF", "#482878FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()
```


####- fitzjohn root prior
```{r}
# Calculating rates for higher frequency transitions (new function)
counts.LF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Larva->Flex", uncert = "confint")
counts.FL_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Flex->Larva", uncert = "confint")
counts.PF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Pupa->Flex", uncert = "confint")
counts.PL_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Pupa->Larva", uncert = "confint")
counts.FP_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Flex->Pupa", uncert = "confint")
counts.FA_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Flex->Adult", uncert = "confint")


# Figure 5a: Plotting rate through time for higher frequency shifts
pdf("newRate_FiveStates_fitz.pdf", height=7, width=7)

rateplot_ey(rates=counts.LF_fitz, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.1, lty=1, color = "#440154FF", confint = T)
         
par(new=T)
segplot_ey(rates=counts.FL_fitz, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.PF_fitz, ylim=c(0,0.02), color= "#2A788EFF" ,  spline=T, width=3, ax=F, lty=1, confint = T)

par(new=T)
segplot_ey(rates=counts.PL_fitz, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty = 2, confint = T)

par(new=T)
segplot_ey(rates=counts.FP_fitz, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FA_fitz, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("Larva->Flex",  "Flex->Larva", "Pupa->Flex", "Pupa->Larva", "Flex->Pupa","Flex->Adult"),  
       col = c("#440154FF", "#482878FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()
```






