---
title: "Counting number of transitions  & rate through time"
author: "Sridhar Halali"
date: "2023-10-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Importing required libraries
```{r setup, message=F}
library(phytools)
library(ape)
library(tidyverse)
library(conflicted)
library(viridis)
```

# Setting working directory
```{r setup, message=F}
setwd("")

```

# Reading stochastic maps saved as .RDS files
```{r}
# Reading stochastic maps (RDS files) if already not in the R environment
SimmapFiveState_flat <- readRDS("simmap_5state_1000sim_ARD_flat.RDS")
SimmapThreeState_flat <- readRDS("simmap_3state_1000sim_ARD_flat.RDS")
SimmapBinaryState_flat <- readRDS("simmap_Binary_1000sim_ARD_flat.RDS")

SimmapFiveState_fitz <- readRDS("simmap_5state_1000sim_ARD_fitzjohn.RDS")
SimmapThreeState_fitz <- readRDS("simmap_3state_1000sim_ARD_fitzjohn.RDS")
SimmapBinaryState_fitz <- readRDS("simmap_Binary_1000sim_ARD_fitzjohn.RDS")

```


# Counting number of transitions

##* Extracting number of transitions on each stochastic maps using countSimmap function
```{r}
# Flat prior
counts_FiveStates_flat <- data.frame(countSimmap(SimmapFiveState_flat)[[1]])
counts_ThreeStates_flat <- data.frame(countSimmap(SimmapThreeState_flat)[[1]])
counts_BinaryStates_flat <- data.frame(countSimmap(SimmapBinaryState_flat)[[1]])

# Fitzjohn prior
counts_FiveStates_fitz <- data.frame(countSimmap(SimmapFiveState_fitz)[[1]])
counts_ThreeStates_fitz <- data.frame(countSimmap(SimmapThreeState_fitz)[[1]])
counts_BinaryStates_fitz <- data.frame(countSimmap(SimmapBinaryState_fitz)[[1]])

```

##* Getting data into shape for plotting
```{r}
head(counts_FiveStates_fitz)

# Replacing '.' from the column names to -> & creating a new column showing root_prior

# Flat prior
names(counts_FiveStates_flat) <- gsub(x = names(counts_FiveStates_flat), pattern = "\\.", replacement = "->")  
counts_FiveStates_flat <- counts_FiveStates_flat %>% mutate(root_prior = "flat")

names(counts_ThreeStates_flat) <- gsub(x = names(counts_ThreeStates_flat), pattern = "\\.", replacement = "->")  
counts_ThreeStates_flat <- counts_ThreeStates_flat %>% mutate(root_prior = "flat")

names(counts_BinaryStates_flat) <- gsub(x = names(counts_BinaryStates_flat), pattern = "\\.", replacement = "->")  
counts_BinaryStates_flat <- counts_BinaryStates_flat %>% mutate(root_prior = "flat")


# Fitzjohn prior
names(counts_FiveStates_fitz) <- gsub(x = names(counts_FiveStates_fitz), pattern = "\\.", replacement = "->")  
counts_FiveStates_fitz <- counts_FiveStates_fitz %>% mutate(root_prior = "fitzjohn")

names(counts_ThreeStates_fitz) <- gsub(x = names(counts_ThreeStates_fitz), pattern = "\\.", replacement = "->")  
counts_ThreeStates_fitz <- counts_ThreeStates_fitz %>% mutate(root_prior = "fitzjohn")

names(counts_BinaryStates_fitz) <- gsub(x = names(counts_BinaryStates_fitz), pattern = "\\.", replacement = "->")  
counts_BinaryStates_fitz <- counts_BinaryStates_fitz %>% mutate(root_prior = "fitzjohn")


#============= Combining dataframes according to state classifications==================================#
# Five states
counts_FiveStates_final <- rbind(counts_FiveStates_flat, counts_FiveStates_fitz)

# Three states
counts_ThreeStates_final <- rbind(counts_ThreeStates_flat, counts_ThreeStates_fitz)

# Binary state
counts_BinaryStates_final <- rbind(counts_BinaryStates_flat, counts_BinaryStates_fitz)
#==================================================================================================#


#================= Converting wide to long data frame ==================================#
# Five states
counts_FiveStates_final <- counts_FiveStates_final %>% 
  select(-N) %>% 
  pivot_longer(!root_prior, names_to = "transitions", values_to = "count") %>% 
  mutate(transitions= factor(transitions),
         root_prior = factor(root_prior))

# Three states
counts_ThreeStates_final <- counts_ThreeStates_final %>% 
  select(-N) %>% 
  pivot_longer(!root_prior, names_to = "transitions", values_to = "count") %>% 
  mutate(transitions= factor(transitions),
         root_prior = factor(root_prior))

# Binary states
counts_BinaryStates_final <- counts_BinaryStates_final %>% 
  select(-N) %>% 
  pivot_longer(!root_prior, names_to = "transitions", values_to = "count") %>% 
  mutate(transitions= factor(transitions),
         root_prior = factor(root_prior))


#================= Removing same transitions ==================================#
counts_FiveStates_final <- counts_FiveStates_final %>% 
  dplyr::filter(!transitions %in% c("Adult->Adult", "Egg->Egg", "Larva->Larva", "Pupa->Pupa", "no_diapause->no_diapause"))

counts_ThreeStates_final <- counts_ThreeStates_final %>% 
  dplyr::filter(!transitions %in% c("Adult->Adult", "Juvenile->Juvenile", "no_diapause->no_diapause"))

counts_BinaryStates_final <- counts_BinaryStates_final %>% 
  dplyr::filter(!transitions %in% c("Diapause->Diapause", "no_diapause->no_diapause"))

```


##* Calculating average number of transitions
```{r}
# Five state
counts_FiveStates_average <- counts_FiveStates_final %>% 
  group_by(root_prior, transitions) %>% 
  summarise(mean_counts = mean(count))


# Three state
counts_ThreeStates_average <- counts_ThreeStates_final %>% 
  group_by(root_prior, transitions) %>% 
  summarise(mean_counts = mean(count))

# Binary state
counts_BinaryStates_average <- counts_BinaryStates_final %>% 
  group_by(root_prior, transitions) %>% 
  summarise(mean_counts = mean(count))

```


##* Plotting number of transitions
```{r}
# Setting the theme
theme_counts <- theme_bw() + 
  theme(panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank())


# Figure 2a: Five state
pdf("transitions_FiveState.pdf", height=6, width=6)

counts_FiveStates_final %>% 
  ggplot(aes(x=count, y= reorder(transitions, count))) +
  
  geom_jitter(aes(fill=root_prior), height=0.2, alpha=0.1, shape=21, stroke=NA) +
  
  geom_jitter(data=counts_FiveStates_average,
             aes(x=mean_counts, y= transitions, fill=root_prior), 
             height = 0.15, shape=21, stroke=0.5, size=4) +
 
  scale_x_continuous(limits= c(-0.9, 250)) +
             
  scale_fill_manual(values = c("flat" = "grey", "fitzjohn" = "salmon")) +

  xlab("Number of transitions") + ylab("Transition between states") +
  theme_counts 

dev.off()


# Figure 2b: Three state
pdf("transitions_ThreeState.pdf", height=6, width=6)

counts_ThreeStates_final %>% 
  ggplot(aes(x=count, y= reorder(transitions, count))) +
  
  geom_jitter(aes(fill=root_prior), height=0.1, alpha=0.1, shape=21, stroke=NA) +
  
  geom_jitter(data=counts_ThreeStates_average,
             aes(x=mean_counts, y= transitions, fill=root_prior), 
             height = 0.09, shape=21, stroke=0.5, size=4) +
  
  scale_x_continuous(limits=c(-0.9, 200)) +
             
  scale_fill_manual(values = c("flat" = "grey", "fitzjohn" = "salmon")) +
  
  xlab("Number of transitions") + ylab("Transition between states") +
  theme_counts 

dev.off()


# Figure 2c: Binary state
pdf("transitions_BinaryState.pdf", height=6, width=6)

counts_BinaryStates_final %>% 
  ggplot(aes(x=count, y= reorder(transitions, count))) +
  
  geom_jitter(aes(fill=root_prior), height=0.07, alpha=0.1, shape=21, stroke=NA) +
  
  geom_jitter(data=counts_BinaryStates_average,
             aes(x=mean_counts, y= transitions, fill=root_prior), 
             height = 0.05, shape=21, stroke=0.5, size=4) +
             
  scale_fill_manual(values = c("flat" = "grey", "fitzjohn" = "salmon")) +
  
  xlab("Number of transitions") + ylab("Transition between states") +
  theme_counts 

dev.off()

```


# Calculating Rate Trough time
Calculating rate through time using 1000 stochastic maps. The rate is calculated by dividing the number transitions by the edge length in a given bin or segment. Here the phylogeny is divided into 30 bins (i.e. 107.60/30), thus, one bin equals to ~3.58 million years. 

##* Loading functions for calculating and plotting rate through time. 
Only some functions are modified from the original code that can be found here (https://github.com/jakeberv/mammal_arboreality). The original getrates(), rateplot() and segplot() functions were especially modified to able to compute and plot confidence intervals for the rates. Please load both source codes. 

```{r}
# Loading original function (can be found here: https://github.com/jakeberv/mammal_arboreality)
# Original article can be found here:  https://doi.org/10.1002/ece3.8114
source("source_rate_through_time_Hughes_et_al_2021.R")

# Loading modified function (based on the original function mentioned above) for incorporating confidence bounds on rate through time calculations 
source("getRates_modified_Etka_Yapar.R")

# Total age of the tree
ButterflyPhylo <- read.nexus("butterfly_phylogeny_pruned.trees")
max(nodeHeights(ButterflyPhylo)) # 107.60

```

##* Binary state diapause clasification

####- flat root prior
```{r}
# Calculating rates for each transition
counts.DF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_flat, type="Diapause->no_diapause", uncert= "confint")

counts.FD_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_flat, type="no_diapause->Diapause", uncert = "confint")


# Figure S7: Plotting rate through time for transitions for binary state diapause classification with flat root prior
pdf("newRate_BinaryState_flat.pdf", height=7, width=7)

rateplot_ey(rates=counts.DF_flat, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint = T)
par(new=T)
segplot_ey(rates=counts.FD_flat, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)
legend("topleft", 
       legend=c("diapause -> no diapause", "no diapause -> diapause"),     
       col = c("#440154FF", "#7AD151FF"), bty = 'n', lty=1, lwd=5)
       
dev.off()

```


####- fitzjohn root prior
```{r}
# Calculating rates for each transition
counts.DF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_fitz, type="Diapause->no_diapause", uncert="confint")
counts.FD_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapBinaryState_fitz, type="no_diapause->Diapause", uncert="confint")


# Figure 5c: Plotting rate through time for transitions for binary state diapause classification with fitzjohn root prior
pdf("newRate_BinaryState_fitz.pdf", height=7, width=7)

rateplot_ey(rates=counts.DF_fitz, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint = T)
par(new=T)
segplot_ey(rates=counts.FD_fitz, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)
legend("topleft", 
       legend=c("diapause -> no diapause", "no diapause -> diapause"),     
       col = c("#440154FF", "#7AD151FF"), bty = 'n', lty=1, lwd=5)
      
dev.off()

```


##* Three state dipauase classification

####- flat root prior 
```{r}
# Calculating rates for each transition
counts.AJ_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Adult->Juvenile", uncert = "confint")
counts.JA_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Juvenile->Adult", uncert = "confint")
counts.JF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Juvenile->no_diapause", uncert = "confint")
counts.FJ_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="no_diapause->Juvenile", uncert = "confint")
counts.FA_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="no_diapause->Adult", uncert = "confint")
counts.AF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_flat, type="Adult->no_diapause", uncert = "confint")

# Figure S7: Plotting rates
pdf("newRate_ThreeState_flat_ColorUpdate.pdf", height=7, width=7)

rateplot_ey(rates=counts.JF_flat, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint=T)

par(new=T)
#segplot_ey(rates=counts.AJ_flat, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint=T)
segplot_ey(rates=counts.AJ_flat, ylim=c(0,0.02), color= "salmon",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.JA_flat, ylim=c(0,0.02), color= "#2A788EFF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.FJ_flat, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.FA_flat, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint=T)

par(new=T)
segplot_ey(rates=counts.AF_flat, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=1, confint=T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("juvenile -> no_diapause", "adult -> juvenile",  "juvenile -> adult", 
                "no diapause -> juvenile", "no diapause -> adult", "adult -> no diapause"),
       #col = c("#482878FF", "#414487FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       col = c("#482878FF", "salmon", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()

```


####- fitzohn root prior
```{r}
# Calculating rates for each transition
counts.AJ_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Adult->Juvenile", uncert = "confint")
counts.JA_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Juvenile->Adult", uncert = "confint")
counts.JF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Juvenile->no_diapause", uncert = "confint")
counts.FJ_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="no_diapause->Juvenile", uncert = "confint")
counts.FA_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="no_diapause->Adult", uncert = "confint")
counts.AF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapThreeState_fitz, type="Adult->no_diapause", uncert = "confint")

# Figure 5b: Plot rate through time
pdf("newRate_ThreeState_fitz_ColorUpdate.pdf", height=7, width=7)

rateplot_ey(rates=counts.JF_fitz, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.2, lty=1, color = "#440154FF", confint = T)

par(new=T)
#segplot_ey(rates=counts.AJ_fitz, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint = T)
segplot_ey(rates=counts.AJ_fitz, ylim=c(0,0.02), color= "salmon",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.JA_fitz, ylim=c(0,0.02), color= "#2A788EFF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FJ_fitz, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FA_fitz, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.AF_fitz, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=1, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("juvenile -> no_diapause", "adult -> juvenile",  "juvenile -> adult", 
                "no diapause -> juvenile", "no diapause -> adult", "adult -> no diapause"),
       #col = c("#482878FF", "#414487FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       col = c("#482878FF", "salmon", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()

```

##* Five state diapause classification
Note that for five state classification, rates are for only the high frequency transitions are calculated (see Figure 2a in the main text for number of transitions between states)

####- flat root prior
```{r}
# Calculating rates only for higher frequency transitions
counts.LF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Larva->no_diapause", uncert = "confint")
counts.FL_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="no_diapause->Larva", uncert = "confint")
counts.PF_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Pupa->no_diapause", uncert = "confint")
counts.PL_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="Pupa->Larva", uncert = "confint")
counts.FP_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="no_diapause->Pupa", uncert = "confint")
counts.FA_flat <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_flat, type="no_diapause->Adult", uncert = "confint")


# Figure S7: Plotting rate through time for higher frequency shifts
pdf("newRate_FiveStates_flat_ColorUpdate.pdf", height=7, width=7)

rateplot_ey(rates=counts.LF_flat, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.1, lty=1, color = "#440154FF", confint = T)
         
par(new=T)
#segplot_ey(rates=counts.FL_flat, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint = T)
segplot_ey(rates=counts.FL_flat, ylim=c(0,0.02), color= "salmon",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.PF_flat, ylim=c(0,0.02), color= "#2A788EFF" ,  spline=T, width=3, ax=F, lty=1, confint = T)

par(new=T)
segplot_ey(rates=counts.PL_flat, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty = 2, confint = T)

par(new=T)
segplot_ey(rates=counts.FP_flat, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FA_flat, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("larva -> no diapause",  "no diapause -> larva", "pupa -> no diapause", 
                "pupa -> larva", "no diapause -> pupa", "no diapause -> adult"),  
       #col = c("#440154FF", "#482878FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       col = c("#440154FF", "salmon", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()

```


####- fitzjohn root prior
```{r}
# Calculating rates for higher frequency shifts
counts.LF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Larva->no_diapause", uncert = "confint")
counts.FL_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="no_diapause->Larva", uncert = "confint")
counts.PF_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Pupa->no_diapause", uncert = "confint")
counts.PL_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="Pupa->Larva", uncert = "confint")
counts.FP_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="no_diapause->Pupa", uncert = "confint")
counts.FA_fitz <- getrates_ey(tree=ButterflyPhylo, bins=30, maps=SimmapFiveState_fitz, type="no_diapause->Adult", uncert = "confint")


# Figure 5a: Plotting rate through time for higher frequency shifts
pdf("newRate_FiveStates_fitz_ColorUpdate.pdf", height=7, width=7)

rateplot_ey(rates=counts.LF_fitz, tree=ButterflyPhylo, ylim=c(0,0.02), spline=T, width=3, alpha=0.1, lty=1, color = "#440154FF", confint = T)
         
par(new=T)
#segplot_ey(rates=counts.FL_fitz, ylim=c(0,0.02), color= "#414487FF",  spline=T, width=3, ax=F, lty=2, confint = T)
segplot_ey(rates=counts.FL_fitz, ylim=c(0,0.02), color= "salmon",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.PF_fitz, ylim=c(0,0.02), color= "#2A788EFF" ,  spline=T, width=3, ax=F, lty=1, confint = T)

par(new=T)
segplot_ey(rates=counts.PL_fitz, ylim=c(0,0.02), color= "#22A884FF",  spline=T, width=3, ax=F, lty = 2, confint = T)

par(new=T)
segplot_ey(rates=counts.FP_fitz, ylim=c(0,0.02), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2, confint = T)

par(new=T)
segplot_ey(rates=counts.FA_fitz, ylim=c(0,0.02), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=2, confint = T)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("larva -> no diapause",  "no diapause -> larva", "pupa -> no diapause", 
                "pupa -> larva", "no diapause -> pupa", "no diapause -> adult"),  
       #col = c("#440154FF", "#482878FF", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       col = c("#440154FF", "salmon", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()

```


# Calculating rates across posterior trees

##* Importing 200 pruned posterior trees
```{r}
# Import pruned multiphylo from Chazot et al. (2019)
ButterflyMultiPruned <- read.nexus("ButterflyPhyloMulti_pruned.trees")


# Calculating maximum node height across trees
max_node_height <- function(tree) {
  node_heights <- nodeHeights(tree)
  return(max(node_heights))
}

# Apply the function to each tree in the set of posterior trees
max_node_heights <- sapply(ButterflyMultiPruned, max_node_height)
max_node_heights <- data.frame(max_node_heights)
max_node_heights

# Figure S11
png("variation_max_nodeheight.png", height=6, width=6, units='in', res=300)
max_node_heights %>% 
  ggplot(aes(x=max_node_heights)) + 
  geom_histogram() +
  theme_bw()
dev.off()

```


##* Preparing data for stochastic mapping 
```{r}
# Importing processed diapause data & pruned phylogeny (see 'preprocessing data & some plots' script for pre-processing steps)
diapause_processed_dat <- read.xlsx("diapause_classification_data_PROCESSED.xlsx")

# Assigning first column as a row names
dat_diapause <- data.frame(diapause_processed_dat, row.names = 1)

# Factorising all columns
dat_diapause <- dat_diapause %>% mutate_if(is.character,as.factor)

# Extracting diapause classification columns and assigning to rownames
DiapauseFive <- setNames(dat_diapause$diapause_five_state, rownames(dat_diapause)) 
DiapauseThree <- setNames(dat_diapause$diapause_three_state, rownames(dat_diapause)) 
DiapauseBinary <- setNames(dat_diapause$diapause_binary, rownames(dat_diapause)) 

```

##* Fitting make.simmap generating 100 stochastic maps per 200 posterior trees only for ARD model with fittzjohn prior
```{r}
# Parallelising make.simmap for running five states as fitting this model takes a lot of time
library(parallel)

detectCores()
cl <- makeCluster(14)
clusterExport(cl=cl, varlist=c("make.simmap", "DiapauseFive"))

par_simmap_FiveStates_fitz <- parLapply(
  cl=cl, 
  ButterflyMultiPruned, 
  function(tree){
    result<-make.simmap(tree = tree, x=DiapauseFive, model="ARD", nsim=100, pi="fitzjohn")
  })

saveRDS(par_simmap_FiveStates_fitz, file="FiveState_200Trees_100sim_ARD_fitz.RDS")


# Three state
simmap_ThreeState_fitzjohn <- make.simmap(ButterflyMultiPruned, DiapauseThree, model= "ARD", 
                                         nsim= 100, pi= "fitzjohn")
saveRDS(simmap_ThreeState_fitzjohn, file= "simmap_3state_100sim_ARD_fitzjohn.RDS")


# Binary state
simmap_Binary_fitzjohn <- make.simmap(ButterflyMultiPruned, DiapauseBinary, model= "ARD", 
                                          nsim= 100, pi= "fitzjohn")
saveRDS(simmap_Binary_fitzjohn, file= "simmap_Binary_200Tree_100Sim_ARD_fitzjohn.RDS")

```


##* Calculating Rate Trough time across posterior trees (20000 stochastic maps in total)
The procedure is same as above but using looped version of the functions (available here: https://github.com/jakeberv/mammal_arboreality). 

###- Binary state diapause clasification 
```{r}
# Reading stochastic maps ran on posterior trees
#SimmapBinaryState_fitz_multi <- readRDS("simmap_Binary_200Tree_100Sim_ARD_fitzjohn.RDS")

# grouping stochastic based on posterior tree identity (splitter function is imported from Hughes et al. (2021) source code
RatesSplitter_binary <- splitter(SimmapBinaryState_fitz_multi, length = 100)
RatesSplitter_binary


# Calculating rates for each transition across posterior trees
counts.DF_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_binary, type="Diapause->no_diapause", log=F)

counts.FD_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_binary, type="no_diapause->Diapause", log=F)


# Figure S12C
# plotting rates for posterior trees
png("Binary_Rates_posterior.png", height=7, width=7, res=300, units='in')
# Plotting rates for all posterior trees
segplotloop(counts.DF_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#440154FF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.FD_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#7AD151FF", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)


# Overlaying rates calculated for single consensus tree
par(new=T)
segplot(rates=counts.DF_fitz, ylim=c(0,0.03), color= "#440154FF",  spline=T, width=3, ax=F, lty=1)
par(new=T)
segplot(rates=counts.FD_fitz, ylim=c(0,0.03), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2)

abline(v = 35, col = "grey", lty = 2, lwd=2)
legend("topleft",
       legend=c("diapause -> no diapause", "no diapause -> diapause"),
       col = c("#440154FF", "#7AD151FF"), bty = 'n', lty=1, lwd=5)
      
dev.off()
```

###- Three state diapause clasification
```{r}
# Reading stochastic maps ran on posterior trees
#SimmapThreeState_fitz_multi <- readRDS("simmap_3state_200Tree_100Sim_ARD_fitzjohn.RDS")


# grouping stochastic based on posterior tree identity
RatesSplitter_three <- splitter(SimmapThreeState_fitz_multi, length = 100)
head(RatesSplitter_three)


# Calculating rates
counts.AJ_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_three, type="Adult->Juvenile", log=F)

counts.JA_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_three, type="Juvenile->Adult", log=F)

counts.JF_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_three, type="Juvenile->no_diapause", log=F)

counts.FJ_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_three, type="no_diapause->Juvenile", log=F)

counts.FA_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_three, type="no_diapause->Adult", log=F)

counts.AF_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = RatesSplitter_three, type="Adult->no_diapause", log=F)


# Figure S12B
# plotting rates for posterior trees
png("3States_Rates_posterior.png", height=7, width=7, res=300, units='in')

segplotloop(counts.JF_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#440154FF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.AJ_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("salmon", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)

segplotloop(counts.JA_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#2A788EFF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.FJ_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#22A884FF", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)

segplotloop(counts.FA_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#7AD151FF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.AF_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#FDE725FF", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)


# Overlaying rates calculated for single consensus tree
par(new=T)
segplot(rates=counts.JF_fitz, ylim = c(0,0.03), color= "#440154FF",  spline=T, width=3, ax=F, lty=1)

par(new=T)
segplot(rates=counts.AJ_fitz, ylim = c(0,0.03), color= "salmon",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.JA_fitz, ylim = c(0,0.03), color= "#2A788EFF",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.FJ_fitz, ylim = c(0,0.03), color= "#22A884FF",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.FA_fitz, ylim = c(0,0.03), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.AF_fitz, ylim = c(0,0.03), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=1)

abline(v = 35, col = "grey", lty = 2, lwd=2)

legend("topleft", 
       legend=c("juvenile -> no diapause", 
                "adult -> juvenile",  
                "juvenile -> adult",
                "no diapause -> juvenile",
                "no diapause -> adult",
                "adult -> no diapause"),
       col = c("#440154FF", "salmon", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()

```

###- Five state diapause clasification
```{r}
# Reading stochastic maps ran on posterior trees
#SimmapThreeState_fitz_multi <- readRDS("simmap_3state_200Tree_100Sim_ARD_fitzjohn.RDS")


# applying splitter() function is not needed as the object is already in the required format (because the make.simmap for five states is ran differently, that is, by parallelizing the run)

# Calculating rates
counts.LF_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = SimmapFiveState_fitz_multi, type= "Larva->no_diapause", log=F)

counts.FL_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = SimmapFiveState_fitz_multi, type= "no_diapause->Larva", log=F)

counts.PF_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = SimmapFiveState_fitz_multi, type= "Pupa->no_diapause", log=F)

counts.PL_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = SimmapFiveState_fitz_multi, type= "Pupa->Larva", log=F)

counts.FP_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = SimmapFiveState_fitz_multi, type= "no_diapause->Pupa", log=F)

counts.FA_fitz_loop <- getratesloop(ButterflyMultiPruned, bins = 30, 
                          splitmaps = SimmapFiveState_fitz_multi, type= "no_diapause->Adult", log=F)


# Figure S12A
# plotting rates for posterior trees

png("5States_Rates_posterior.png", height=7, width=7, res=300, units='in')

segplotloop(counts.LF_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#440154FF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.FL_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("salmon", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)

segplotloop(counts.PF_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#2A788EFF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.PL_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#22A884FF", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)

segplotloop(counts.FP_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#7AD151FF", 0.2),
            spline=T, width=0.2, ax=T, lty = 1)

segplotloop(counts.FA_fitz_loop, ylim = c(0,0.03), xlim= c(110, 0), color = alpha("#FDE725FF", 0.2),
            spline=T, width=0.5, ax=T, lty = 1)


# Overlaying rates calculated for single consensus tree
par(new=T)
segplot(rates=counts.LF_fitz, ylim = c(0,0.03), color= "#440154FF",  spline=T, width=3, ax=F, lty=1)

par(new=T)
segplot(rates=counts.FL_fitz, ylim = c(0,0.03), color= "salmon",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.PF_fitz, ylim = c(0,0.03), color= "#2A788EFF",  spline=T, width=3, ax=F, lty=1)

par(new=T)
segplot(rates=counts.PL_fitz, ylim = c(0,0.03), color= "#22A884FF",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.FP_fitz, ylim = c(0,0.03), color= "#7AD151FF",  spline=T, width=3, ax=F, lty=2)

par(new=T)
segplot(rates=counts.FA_fitz, ylim = c(0,0.03), color= "#FDE725FF",  spline=T, width=3, ax=F, lty=1)

abline(v = 35, col = "grey", lty = 2, lwd=2)


legend("topleft", 
       legend=c("larva -> no diapause",  
                "no diapause -> larva", 
                "pupa -> no diapause", 
                "pupa -> larva", 
                "no diapause -> pupa", 
                "no diapause -> adult"),  
       col = c("#440154FF", "salmon", "#2A788EFF", "#22A884FF", "#7AD151FF", "#FDE725FF"),
       bty = 'n', lty=1, lwd=4)

dev.off()

```


