---
title: "plotting_uncertainties_recons"
author: "Sridhar Halali & Etka Yapar"
date: "2023-12-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Importing libraries
```{r, message=FALSE}
library(phytools)
library(ape)

# Setting working directory
setwd("")

# Importing the pruned tree
tree <- read.nexus("butterfly_phylogeny_pruned.trees")
tree

```


A simple metric to highlight nodes which have ambiguous reconstructions across ancestral reconstructions methods. I am using three different reconstruction methods (max likelihood, stochastic mapping and hidden rates models) across flat and fitzjohn root prior to see how much reconstructions differ. The general observation was the some nodes show same patterns while some nodes (especially at the deeper nodes) are highly uncertainties across methods/root priors. So we needed a simple metric to summarise these uncertainties in the ancestral state estimation for the same node across methods/root priors. We did this by calculating total sum of squares using kmeans clustering. The higher the sum of squares, higher the uncertainty and vice versa


# Visualizing uncertainty for five state classification

##* function for summarising uncertainty for ancestral estimations using five state classification (function written by Etka Yapar)
```{r}
plot_uncertain_recons_5 <- function(recons_obj_names, phy){ 
  # recons_obj_names = 'ace' element from different reconstructions
  
  Nnodes = Nnode(phy) # total number of nodes in the phylogeny
  
  calc_var <- function(df){
    # N X 2 df/matrix where N is number of states and columns are two recons.
    df <- t(df) # transpose the dataframe
    totss <- kmeans(df, 2)$totss # getting total sum of squared, 2 arbitrary clusters are used
    return(totss)
  }
  
  # function to apply calc_var function independently for all nodes 
  six_rec_vars <- sapply(1:Nnodes, function(node) {
    this_node <- sapply(recons_objs_five, function(rec){
      this_recons <- get(rec)
      return(this_recons[node,])
    })
    return(calc_var(this_node))
  })
  
  # creating a gradient of colors, more redder = more uncertain nodes
  cscale <- scales::gradient_n_pal(c("white","red"),c(min(six_rec_vars), max(six_rec_vars)))
  cols = sapply(six_rec_vars, function(x){cscale(x)})
  leg_cols = sapply(seq(min(six_rec_vars), max(six_rec_vars), length.out=1000), cscale)
  
  # plotting nodes on phylogeny
  plotTree(phy, ftype='off', type='fan')
  nodelabels(pch=21, bg=cols, col='black',cex=1.2)
  add.color.bar(x= -110, y= 100, leg=50, cols=leg_cols, lims=round(range(six_rec_vars),3), 
                title="Tot. Sum.Sq.", subtitle="", prompt=F)
}
```


##* Importing ancestral estimation objects as .RDS files using different methods (maximum likelihood, stochastic mapping, hidden rates)
```{r}
#=========== Importing Maximum likelihood estimation of transition rates (from phytools) ====================#
five_fitz_ML <- readRDS("fitMK_ARD_Diapausefive_fitz.RDS")
five_flat_ML <- readRDS("fitMK_ARD_Diapausefive_flat.RDS")

# Computing marginal state ancestral estimation
anc_five_flat <- ancr(five_flat_ML)$ace
anc_five_fitz <- ancr(five_fitz_ML)$ace


#=========== Importing summary of stochastic maps object (from phytools) ====================#
summary_five_flat <- readRDS("summary_5state_ARD_flat.RDS")
summary_five_fitz <- readRDS("summary_5state_ARD_fitz.RDS")

# Extracting marginal state ancestral states
simmap_five_flat <- summary_five_flat$ace
simmap_five_fitz <- summary_five_fitz$ace


#=============== Importing hidden rates objects (from corHMM) =============================#
hmm_five_flat <- readRDS("corhmm_5State_Umbral_hrm_flat.RDS")
hmm_five_fitz <- readRDS("corhmm_5State_Umbral_hrm_fitz.RDS")

# Slightly modified function to hide hidden states (see here: )
hide_hidden<-function(object,...){
  ss<-colnames(object) # getting column names of ancr object
  ss<-ss[-grep("*",ss,fixed=TRUE)] # removing columns containing '*'
  
  # Makes an empty matrix corresponding to column and row names of the ancr object
  anc<-matrix(0,nrow(object),length(ss), dimnames=list(rownames(object),ss)) 
  
  # Loop to get rowsums   
  for(i in 1:length(ss)){
    anc[,ss[i]]<-rowSums(object[,grep(ss[i],
                                          colnames(object))])
  }
  anc
}

# Extracting states and prepping the data frame (by renaming column names)

# Five states- flat prior
hmm_five_flat_states <- hmm_five_flat$states 
colnames(hmm_five_flat_states) <- c("Adult", "Egg", "Larva", "no_diapause", "Pupa", 
                                    "Adult*", "Egg*", "Larva*", "no_diapause*", "Pupa*")

# Five states-fitz prior
hmm_five_fitz_states <- hmm_five_fitz$states 
colnames(hmm_five_fitz_states) <- c("Adult", "Egg", "Larva", "no_diapause", "Pupa", 
                                    "Adult*", "Egg*", "Larva*", "no_diapause*", "Pupa*")

# Applying the hide_hidden function
anc_five_hmm_flat <- hide_hidden(hmm_five_flat_states)
anc_five_hmm_fitz <- hide_hidden(hmm_five_fitz_states)

```


##* Shaping the data before fitting the function**
```{r}
# list of the ancestral reconstructions
recons_objs_five <- grep("^hmm", grep("summary", grep("_five", ls(), val=T), 
                                      invert=T, val=T), perl=T, invert=T, val=T)
recons_objs_five


# The column order (i.e. states) can sometimes be different for different methods which the affects later calculations. So first sorting columns in alphabetical orders across reconstructions
col_order <- sort(colnames(anc_five_fitz))
col_order

this_node <- sapply(recons_objs_five, function(rec){
  this_recons <- get(rec)
  this_recons <- this_recons[, col_order]
  assign(rec, this_recons, envir=.GlobalEnv)
  return(NULL)
})

this_node

this_node <- sapply(recons_objs_five, function(rec){
  this_recons <- get(rec)
  print(colnames(this_recons))
  return(NULL)
})


# Figure S6: Plotting uncertainty
pdf("uncertain_five_state.pdf", height=8, width = 8)
plot_uncertain_recons_5(recons_objs_five, tree)
dev.off()

```


# Visualizing uncertainty for three state classification
Repeating same steps as before
```{r}
#=========== Importing Maximum likelihood estimation of transition rates (from phytools) ====================#
# importing RDS files
three_fitz <- readRDS("fitMK_ARD_Diapausethree_fitz.RDS")
three_flat <- readRDS("fitMK_ARD_Diapausethree_flat.RDS")

# Computing marginal state ancestral estimation
anc_three_flat <- ancr(three_flat)$ace
anc_three_fitz <- ancr(three_fitz)$ace

#=========== Importing summary of stochastic maps object (from phytools) ====================#
summary_three_flat <- readRDS("summary_3state_ARD_flat.RDS")
summary_three_fitz <- readRDS("summary_3state_ARD_flat.RDS")

# Extracting marginal state ancestral states
simmap_three_flat <- summary_three_flat$ace
simmap_three_fitz <- summary_three_fitz$ace


#=============== Importing hidden rates objects (from corHMM) =============================#

# importing hidden rate fitted RDS file
hmm_three_flat <- readRDS("corhmm_3State_Umbral_hrm_flat.RDS")
hmm_three_fitz <- readRDS("corhmm_3State_Umbral_hrm_fitz.RDS")


# Extracting states and prepping the dataframe

# Three states- flat prior
hmm_three_flat_states <- hmm_three_flat$states 

colnames(hmm_three_flat_states) <- c("Adult", "Juvenile", "no_diapause",
                                      "Adult*", "Juvenile*", "no_diapause*")
                                    
# Five states-fitz prior
hmm_three_fitz_states <- hmm_three_fitz$states 
colnames(hmm_three_fitz_states) <- c("Adult", "Juvenile", "no_diapause",
                                      "Adult*", "Juvenile*", "no_diapause*")

# Applying the hide_hidden function
anc_three_hmm_flat <- hide_hidden(hmm_three_flat_states)
anc_three_hmm_fitz <- hide_hidden(hmm_three_fitz_states)

```


##* Shaping the data before fitting the function & plotting uncertainty
```{r}
# list of the ancestral reconstructions
recons_objs_three <- grep("^hmm", grep("summary", grep("_three", ls(), val=T), 
                                      invert=T, val=T), perl=T, invert=T, val=T)
recons_objs_three


# the column order (i.e. states) seems to be different for different methods which the affects later calculations. So first sorting columns in alphabetical orders across reconstructions
col_order <- sort(colnames(anc_three_fitz))
col_order

this_node <- sapply(recons_objs_three, function(rec){
  this_recons <- get(rec)
  this_recons <- this_recons[, col_order]
  assign(rec, this_recons, envir=.GlobalEnv)
  return(NULL)
})

```

##* Function for summarising uncertainty for three state classification
```{r}
plot_uncertain_recons_3 <- function(recons_objs_names, phy){ 
  # recons_obj_names = 'ace' element from different reconstructions
  
  Nnodes = Nnode(phy) # total number of nodes in the phylogeny
  
  calc_var <- function(df){
    # N X 2 df/matrix where N is number of states and columns are two recons.
    df <- t(df) # transpose the dataframe
    totss <- kmeans(df, 2)$totss # getting total sum of squared, 2 arbitrary clusters are used
    return(totss)
  }
  
  # function to apply calc_var function independently for all nodes 
  six_rec_vars <- sapply(1:Nnodes, function(node) {
    this_node <- sapply(recons_objs_three, function(rec){
      this_recons <- get(rec)
      return(this_recons[node,])
    })
    return(calc_var(this_node))
  })
  
  # creating a gradient of colors, more redder = more uncertain nodes
  cscale <- scales::gradient_n_pal(c("white","red"),c(min(six_rec_vars), max(six_rec_vars)))
  cols = sapply(six_rec_vars, function(x){cscale(x)})
  leg_cols = sapply(seq(min(six_rec_vars), max(six_rec_vars), length.out=1000), cscale)
  
  # plotting nodes on phylogeny
  plotTree(phy, ftype='off', type='fan')
  nodelabels(pch=21, bg=cols, col='black',cex=1.2)
  add.color.bar(x= -110, y= 100, leg=50, cols=leg_cols, lims=round(range(six_rec_vars),3), 
                title="Tot. Sum.Sq.", subtitle="", prompt=F)
}

# Figure S6: Plotting
pdf("uncertainty_three_state.pdf", height=8, width=8)

plot_uncertain_recons_3(recons_objs_three, tree)

dev.off()

```


# Visualizing uncertainty for two state classification

```{r}
#=========== Importing Maximum likelihood estimation of transition rates (from phytools) ====================#
# Importing .RDS files
binary_fitz <- readRDS("fitMK_ARD_DiapauseBinary_fitz.RDS")
binary_flat <- readRDS("fitMK_ARD_DiapauseBinary_flat.RDS")

# Computing marginal state ancestral estimation
anc_binary_flat <- ancr(binary_flat)$ace
anc_binary_fitz <- ancr(binary_fitz)$ace


#=========== Importing summary of stochastic maps object (from phytools) ====================#
summary_binary_flat <- readRDS("summary_Binary_ARD_flat.RDS")
summary_binary_fitz <- readRDS("summary_Binary_ARD_fitz.RDS")

# Extracting marginal state ancestral states
simmap_binary_flat <- summary_binary_flat$ace
simmap_binary_fitz <- summary_binary_fitz$ace


#=============== Importing hidden rates objects (from corHMM) =============================#

# importing hidden rate fitted RDS file
hmm_binary_flat <- readRDS("corhmm_2State_Umbral_hrm_flat.RDS")
hmm_binary_fitz <- readRDS("corhmm_2State_Umbral_hrm_fitz.RDS")

# Extracting states and prepping the dataframe

# Binary states- flat prior
hmm_binary_flat_states <- hmm_binary_flat$states 
colnames(hmm_binary_flat_states) <- c("Diapause", "no_diapause", "Diapause*", "no_diapause*")

# Binary states-fitz prior
hmm_binary_fitz_states <- hmm_binary_fitz$states 
colnames(hmm_binary_fitz_states) <- c("Diapause", "no_diapause", "Diapause*", "no_diapause*")

# Applying the hide_hidden function
anc_binary_hmm_flat <- hide_hidden(hmm_binary_flat_states)
anc_binary_hmm_fitz <- hide_hidden(hmm_binary_fitz_states)

```


##* Shaping the data before fitting the function
```{r}
# list of the ancestral reconstructions
recons_objs_binary <- grep("^hmm", grep("summary", grep("_binary", ls(), val=T), 
                                      invert=T, val=T), perl=T, invert=T, val=T)
recons_objs_binary


# the column order (i.e. states) seems to be different for different methods which the affects later calculations. So first sorting columns in alphabetical orders across reconstructions
col_order <- sort(colnames(anc_binary_fitz))
col_order

this_node <- sapply(recons_objs_binary, function(rec){
  this_recons <- get(rec)
  this_recons <- this_recons[, col_order]
  assign(rec, this_recons, envir=.GlobalEnv)
  return(NULL)
})

this_node
```

##* Function for summarising uncertainty for three state classification & plotting uncertainty
```{r}
plot_uncertain_recons_2 <- function(recons_objs_names, phy){ 
  # recons_obj_names = 'ace' element from different reconstructions
  
  Nnodes = Nnode(phy) # total number of nodes in the phylogeny
  
  calc_var <- function(df){
    # N X 2 df/matrix where N is number of states and columns are two recons.
    df <- t(df) # transpose the dataframe
    totss <- kmeans(df, 2)$totss # getting total sum of squared, 2 arbitrary clusters are used
    return(totss)
  }
  
  # function to apply calc_var function independently for all nodes 
  six_rec_vars <- sapply(1:Nnodes, function(node) {
    this_node <- sapply(recons_objs_binary, function(rec){
      this_recons <- get(rec)
      return(this_recons[node,])
    })
    return(calc_var(this_node))
  })
  
  # creating a gradient of colors, more redder = more uncertain nodes
  cscale <- scales::gradient_n_pal(c("white","red"),c(min(six_rec_vars), max(six_rec_vars)))
  cols = sapply(six_rec_vars, function(x){cscale(x)})
  leg_cols = sapply(seq(min(six_rec_vars), max(six_rec_vars), length.out=1000), cscale)
  
  # plotting nodes on phylogeny
  plotTree(phy, ftype='off', type='fan')
  nodelabels(pch=21, bg=cols, col='black',cex=1.2)
  add.color.bar(x= -110, y= 100, leg=50, cols=leg_cols, lims=round(range(six_rec_vars),3), 
                title="Tot. Sum.Sq.", subtitle="", prompt=F)
}


# Figure S6: Plotting uncertainty
pdf("uncertain_binary_state.pdf", height=8, width = 8)

plot_uncertain_recons_2(recons_objs_binary, tree)

dev.off()
```
