---
title: "Preprocessing data & making some basic plots"
author: "Sridhar Halali"
date: "2023-10-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Import required libraries & set working directory
```{r setup, message=F}
library(tidyverse)
library(ape)
library(openxlsx)
library(geiger)
library(viridis)
library(conflicted)
library(phytools)

# set working directory
setwd("")

setwd("/Users/sridhar/Library/CloudStorage/OneDrive-LundUniversity/LU box backup/Evolution of diapause/evo of diapause/Updated_diapause_analyses_19_sep_2023/diapause_reanalyses_Mar 2024/GitHub_uploading_codes_EvoLett_22 July 2024")
list.files()
```

# Importing diapause data and doing some preprocessing
```{r}
# Importing the raw data which contains diapause data 
dat_diapause <- read.xlsx("diapause_classification_data_RAW.xlsx")
head(dat_diapause)


# Choosing only required columns
dat_pruned <- dat_diapause %>% 
  select(tree_label, tree_genus, tree_species, family, 
         diapause_5_state_revised, diapause_3_state_revised, diapause_2_state_revised) 
         

# Convert all character columns to factor
dat_pruned <- dat_pruned %>% mutate_if(is.character, as.factor)


# Remove 'skip' & 'skip high altitude' factor from 5, 3 and 2 state diapause  classification and remove from the data
conflicted::conflicts_prefer(dplyr::filter)

dat_pruned_skip <- dat_pruned %>% 
  filter(!diapause_5_state_revised  %in%  c("skip", "skip_high altitude"))

```


# Recoding some diapause states
Some taxa in the data can diapause in both larval and pupal stage. We're recoding these taxa as 'larval' diapause (see Methods in the main for details). Also, recoding none/flexible state as 'Flex' for simplicity. 
Also, for one of the taxa (tip label = UK78_Iliana_sp_) the family is wrongly defined so this will mistake will be corrected in this chunk.

```{r}
dat_recoded <- dat_pruned_skip %>% 
  
  mutate(diapause_five_state = case_when(
    diapause_5_state_revised == "LP" ~ "Larva",
    TRUE~diapause_5_state_revised), 
    
    diapause_three_state = case_when(
      TRUE~diapause_3_state_revised), 
    
    diapause_binary = case_when(
      diapause_2_state_revised == "true_diapause" ~ "Diapause",
      TRUE~diapause_2_state_revised))

# Change the family of this one tree label which is wrongly assigned in the original data. Changing family of "UK78_Iliana_sp_" tree label from Riodinidae to Hesperiidae
row_number <- which(dat_recoded$tree_label == "UK78_Iliana_sp_")

dat_recoded$family[dat_recoded$tree_label=="UK78_Iliana_sp_"] <- "Hesperiidae"
print(dat_recoded[row_number, ])

# Factorising all characters to factors
dat_recoded <- dat_recoded %>% mutate_if(is.character, as.factor)
glimpse(dat_recoded)


# Number of taxa in each family
dat_recoded %>% group_by(family) %>%  count()

# Counting number of taxa in each state
dat_recoded %>% group_by(diapause_five_state) %>%  count()
dat_recoded %>% group_by(diapause_three_state) %>%  count()
dat_recoded %>% group_by(diapause_binary) %>%  count()

dat_recoded %>% group_by(family) %>%  count()

# exporting processed data file
#write.xlsx(dat_recoded, file= "diapause_classification_data_PROCESSED.xlsx")

```


# Making some basic plots
```{r}
# Reordering levels of factors
dat_recoded$diapause_five_state <- factor(dat_recoded$diapause_five_state, 
                                          levels= c("Egg", "Larva", "Pupa", "Adult", "no_diapause"))

dat_recoded$diapause_three_state <- factor(dat_recoded$diapause_three_state, 
                                          levels= c("Juvenile", "Adult", "no_diapause"))

dat_recoded$diapause_binary <- factor(dat_recoded$diapause_binary, 
                                          levels= c("Diapause", "no_diapause"))


#================== Plotting number of taxa in each diapause state across classifications =================================#

# Figure 1a: number of taxa in each state for the five state classification  
pdf("genera_number_five_state.pdf", height=5, width=5)

dat_recoded %>% group_by(diapause_five_state) %>% count() %>%
  ggplot(aes(diapause_five_state, n)) +
  geom_bar(stat= "identity", fill="cadetblue", color="black", lwd=0.2) +
  xlab("") +  ylab("Number of taxa") + 
  theme_bw()
    
dev.off()


# Figure  S2: number of taxa in each state for the three state classification  
pdf("genera_number_three_state.pdf", height=5, width=5)

dat_recoded %>% group_by(diapause_three_state) %>% count() %>%
  ggplot(aes(diapause_three_state, n)) +
  geom_bar(stat= "identity", fill="cadetblue", color="black", lwd=0.2, width=0.5) +
  xlab("") +  ylab("Number of genera") + 
  theme_bw()
  
dev.off()


# Figure  S2: number of taxa in each state for the binary state classification  
pdf("genera_number_binary_state.pdf", height=5, width=5)

dat_recoded %>% group_by(diapause_binary) %>% count() %>%
  ggplot(aes(diapause_binary, n)) +
  geom_bar(stat= "identity", fill="cadetblue", color="black", lwd=0.2, width=0.5) +
  xlab("") +  ylab("Number of genera") + 
  theme_bw()
  
dev.off()


#======= Calculating relative proportion of each diapausing state per family ===============#
# Note that family Hedylidae is removed as this was represented by only two taxa in our phylogenetic tree

# Five state classification
DatPropFive <- dat_recoded %>%
  filter(!family=="Hedylidae") %>% 
  group_by(family, diapause_five_state) %>%
  summarise(n = n()) %>%
  mutate(prop = (n / sum(n))*100)

# Three states
DatPropThree <- dat_recoded %>%
  filter(!family=="Hedylidae") %>% 
  group_by(family, diapause_three_state) %>%
  summarise(n = n()) %>%
  mutate(prop = (n / sum(n))*100)
DatPropThree

# Binary state
DatPropBinary <- dat_recoded %>%
  filter(!family=="Hedylidae") %>% 
  group_by(family, diapause_binary) %>%
  summarise(n = n()) %>%
  mutate(prop = (n / sum(n))*100)

DatPropBinary


# Setting levels for families for plotting
unique(DatPropFive$family)

DatPropFive$family <- factor(DatPropFive$family, 
                             levels= c("Papilionidae", "Pieridae", "Nymphalidae",
                                       "Lycaenidae", "Hesperiidae", "Riodinidae"))
                                       

DatPropThree$family <- factor(DatPropThree$family, 
                             levels= c("Papilionidae", "Pieridae", "Nymphalidae",
                                       "Lycaenidae", "Hesperiidae", "Riodinidae"))

DatPropBinary$family <- factor(DatPropBinary$family, 
                             levels= c("Papilionidae", "Pieridae", "Nymphalidae",
                                       "Lycaenidae", "Hesperiidae", "Riodinidae"))
                                       

# Figure S4: relative proportion of diapause states per family for five state classification
pdf("proportion_5_state.pdf", height=6, width=6)

DatPropFive %>% 
  ggplot(aes(x= family, y= prop, fill= diapause_five_state)) + 
  geom_bar(stat="identity") +
  ylab("Proportion of diapausing strategies") + xlab("") + 
  scale_fill_viridis(discrete = TRUE) +
  geom_text(aes(label=sprintf("%0.1f", prop)), 
            position = position_stack(vjust = 0.5), size = 3)+
  theme_bw() + theme(legend.position="bottom")

dev.off()


# Figure S4: relative proportion of diapause states per family for three state classification
pdf("proportion_3_state.pdf", height=6, width=6)

DatPropThree %>% 
  ggplot(aes(x= family, y= prop, fill= diapause_three_state)) + 
  geom_bar(stat="identity") +
  ylab("Proportion of diapausing strategies") + xlab("") + 
  scale_fill_viridis(discrete = TRUE) +
  geom_text(aes(label=sprintf("%0.1f", prop)), 
            position = position_stack(vjust = 0.5), size = 3)+
  theme_bw() + theme(legend.position="bottom")

dev.off()


# Figure S4: relative proportion of diapause states per family for binary state classification
pdf("proportion_2_state.pdf", height=6, width=6)

DatPropBinary %>% 
  ggplot(aes(x= family, y= prop, fill= diapause_binary)) + 
  geom_bar(stat="identity") +
  ylab("Proportion of diapausing strategies") + xlab("") + 
  scale_fill_viridis(discrete = TRUE) +
  geom_text(aes(label=sprintf("%0.1f", prop)), 
            position = position_stack(vjust = 0.5), size = 3)+
  theme_bw() + theme(legend.position="bottom")

dev.off()

```


# Starting with phylogenetic analyses- first preprocessing the data & phylogeny
```{r}
# Assigning first column as a row names
dat_diapause <- data.frame(dat_recoded, row.names = 1)
head(dat_diapause)
length(dat_diapause$tree_genus)  # 952 tips

# Importing phylogeny from Chazot et al. (2019). Systematic Biology. 68(5), 797-813 
#phylo_genera <- read.nexus("butterfly_phylogeny_Chazot_et_al_2019.tre")

phylo_genera <- read.nexus("S_3b_core_analysis_median_ages.tre")

plotTree(phylo_genera, type="fan", ftype='off')

# Pruning the phylogeny based on the data
phylo_check1 <- name.check(phylo_genera, dat_diapause)
phylo_check1

# Dropping these list of tips which are in the tree but not in the data
ButterflyPhylo <- drop.tip(phylo_genera, phylo_check1$tree_not_data)
ButterflyPhylo

plotTree(ButterflyPhylo, type="fan", ftype= "off")

# exporting pruned phylogeny
#write.nexus(ButterflyPhylo, file= "butterfly_phylogeny_pruned.trees")

```


# Missing species
Investigating how missing species (for which diapause data could not be collected) are distributed across the phylogeny and check the distribution of branch lengths for these species

```{r}
# data frame for missing species
dat_missing_species <- dat_pruned %>% 
  filter(diapause_5_state_revised  %in%  c("skip", "skip_high altitude"))

dat_missing_species

# creating vector of the missing species
missing_species_list <- as.vector(dat_missing_species$tree_label)

# Figure S1: Highlighting the missing species on the phylogeny using add.arrow function from phytools
pdf("missing_species.pdf", height=7, width = 7)
plotTree(phylo_genera, type="fan", ftype='off', color="grey", lwd=1)
add.arrow(phylo_genera, tip= missing_species_list, 
          lwd=1, arrl=5, col="red")
dev.off()


#================ Extracting branch lengths ==================#
# extracting tip names
tips <- phylo_genera$tip.label
tips

# first get the node numbers of the tips
nodes<-sapply(tips,function(x,y) which(y==x),y=phylo_genera$tip.label)
nodes

 # Then get the edge lengths for those nodes
edge.lengths<-setNames(phylo_genera$edge.length[sapply(nodes,
    function(x,y) which(y==x),y=phylo_genera$edge[,2])],names(nodes))

edge.lengths <- data.frame(edge.lengths)
edge.lengths

# rowname to columns
edge.lengths <- edge.lengths %>% rownames_to_column("tip_label")
head(edge.lengths)

 
# subsetting to the list of missing species
missing_species_edges <- edge.lengths %>% 
  filter(tip_label %in% dat_missing_species$tree_label)

missing_species_edges


# Figure S1: plotting edge lenghts of missing species on top of entire distribution of edge lengths
pdf("missing_species_histogram_new.pdf", height=5, width=6)
edge.lengths %>% 
  ggplot(aes(x=edge.lengths)) +
  geom_histogram(bins=60, alpha=0.7, color="black", size=0.05) +
  geom_segment(data = data.frame(x = missing_species_edges$edge.lengths, y = -2, yend = 1), 
                      aes(x = x, xend = x, y = y, yend = yend), 
                      linetype = "solid", color = "red", size = 0.4) +
  xlab("Edge lenghts (in millions of years)") + ylab("Count") +
  theme_bw()
dev.off()

```

